---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
###To help preprocess the data for later analysis, the following packages will be loaded up:

```{r load libraries for later}
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```
  
###Reading in the data:
```{r read in data}
setwd("D:/Classes/Data Science Track (Johns Hopkins)/05 Reproducible Research/Week 2/Peer Assesment 1/RepData_PeerAssessment1")
dat <- read.csv("activity.csv")
dat <- tbl_df(dat)
```
  
###Preprocessing the data to remove the >2,000 observations with NA for steps
```{r remove NA observations}
datClean <- filter(dat, !is.na(steps))
```
  
## What is mean total number of steps taken per day?
    
###Steps taken by day:
```{r group data by day}
steps <-
    datClean %>%
    select(steps:date) %>%
    group_by(date) %>%
    summarize(sum(steps))
        
print.data.frame(steps)
```
  
###Histogram showing the distribution and frequency of the total steps taken each day:
```{r create histogram for total steps taken per day}
qplot(steps$`sum(steps)`,
      data = steps,
      binwidth = 2500,
      xlab = "Total Steps",
      ylab = "Frequency",
      main = "Total Steps Taken Each Day")
```
  
###The mean and median of the total number of steps taken per day:


```{r group data by day DELETE}
stepsSum <-
    steps %>%
    summarize(Mean = mean(steps$`sum(steps)`), Median = median(steps$`sum(steps)`))

print(stepsSum)
```
  
## What is the average daily activity pattern?
  
```{r remove dates, group by interval}
byInterval <-
    datClean %>%
    select(-date) %>%
    group_by(interval) %>%
    summarize(mean = mean(steps)) 

plot(byInterval, type="l",
     xlab="Interval",
     ylab="Number of Steps",
     main="Daily Activity Pattern")
```
  
###Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r find the interval w/ max steps}
byInterval %>%
    filter(mean == max(mean)) %>%
    print
```
  
## Imputing missing values

###Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r filter for NAs}
missingValues <- filter(dat, is.na(steps))

sum(is.na(missingValues$steps))
```
  
###We're going to fill in the missing 2,304 step observations by using the mean number of steps for that interval.  
```{r fill in missing values}
imputedDat <- merge(dat, byInterval)
imputedDat$steps[is.na(imputedDat$steps)] <- imputedDat$mean[is.na(imputedDat$steps)]

```
  
###Histogram of the total number of steps taken each day 
```{r histogram of imputedDat}
imputedDatbyDay <-
    imputedDat %>%
    group_by(date) %>%
    summarize(sum(steps))

qplot(imputedDatbyDay$`sum(steps)`,
      binwidth = 2500,
      xlab = "Total Steps",
      ylab = "Frequency",
      main = "Total Steps Taken Each Day (Imputed Data)")
```
  
###Mean and median of the total number of steps taken per day when using the imputed data:
```{r calc new mean and median}
imputedStepsSum <-
    imputedDatbyDay %>%
    summarize(Mean = mean(imputedDatbyDay$`sum(steps)`), Median = median(imputedDatbyDay$`sum(steps)`)) 

print(imputedStepsSum)
```

###Impact of including imputed data  
```{r calc changes}
meanChange <- stepsSum$Mean - imputedStepsSum$Mean
medianChange <- stepsSum$Median - imputedStepsSum$Median

print(meanChange)
print(medianChange)
```

The mean did not change, although the median decreased by 1.19 steps per day.

## Are there differences in activity patterns between weekdays and weekends?

###Create new factor variable showing whether each day is a weekday or weekend
```{r convert factor to date and add new factor for weekday/weekend}
imputedDatWeekday <- imputedDat
imputedDatWeekday$date <- as.Date(imputedDatWeekday$date)

# imputedDatWeekday %>%
#     mutate(DayOfWeek = ifelse(weekdays(date) == c("Saturday", "Sunday"), "Weekend", "Weekday"))

imputedDatWeekday <- mutate(imputedDatWeekday, DayOfWeek = ifelse(weekdays(date) == c("Saturday", "Sunday"), "Weekend", "Weekday"))
```

###Weekdays vs Weekends  

```{r line graph weekdays vs weekends}
imputedDatWeekdayInterval <-
    imputedDatWeekday %>%
    group_by(DayOfWeek, interval) %>%
    summarize(mean = mean(steps)) 

qplot(interval, 
      mean, 
      data = imputedDatWeekdayInterval, 
      geom = "line") + facet_grid( DayOfWeek ~ .)
```